---
title: "Joins"
author: "K. Shoemaker - DATA 2401"
format: 
  revealjs:
    scrollable: true
    smaller: true
    embed-resources: true
editor: visual
execute:
  echo: true
---

```{r}
#| message: false
#| include: false

library(tidyverse, warn.conflicts = F)
```

## Joins

Sometimes you store data in different dataframes. Consider: you're a charity, and you have one data frame for donations, and one for email and phone number of your donors.

- This helps with data storage: If you store every donation with the email and phone number, you're wasting space.

- This helps with data updating: If you need to update a donor's phone number, you only have to do it in one spot, not on every row of donations they've done.

*Those of you who have worked in SQL, this type of thinking should look familiar.*

------------------------------------------------------------------------

So, how do we do this?

You join data through data that is in both datasets: you need a way to tell the program how to join them.

You identify columns that are in both tables, then use these columns to "match" the corresponding rows.

------------------------------------------------------------------------

So, suppose we have this table of donations:

| Name    | Amount | Date      |
|---------|--------|-----------|
| Sam     | 100    | 2/12/2019 |
| Mike    | 50     | 2/12/2019 |
| Alice   | 100    | 4/23/2019 |
| Mike    | 50     | 5/17/2019 |
| Maria   | 100    | 6/7/2019  |
| Susanne | 50     | 7/2/2019  |

And this table of donors:

| Name    | Email                  |
|---------|------------------------|
| Sam     | sam\@gmail.com         |
| Alice   | aliceemail\@gmail.com  |
| Susanne | susanneroks\@gmail.com |
| Mike    | michael\@hotmail.com   |
| Mary    | mary\@gmail.com        |

We would want to use "Name" to identify which rows correspond to each other in the table.

> Note: This is not actually a great way to store my data... in the real world, there is going to definitely be more than one "Sam" or "Maria", so this wouldn't be a unique identifier

Now, how does R decide which order they go in and what gets left in or left out? You need to specify which type of join you want.

------------------------------------------------------------------------

## Types of Joins

There are two main types of joins, mutating joins and filtering joins.

- Mutating joins add all the columns from one table to the other, combining the variables.

- Filtering joins only take the variables from the first table, but filter the rows based on whether or not that row is present in the second table.

------------------------------------------------------------------------

### Mutating Joins

The first kind of join combines variables from both data frames.

Straight from the R documentation:

- `left_join(x, y, by = )`: return all rows from x, and all columns from x and y. Rows in x with no match in y will have NA values in the new columns. If there are multiple matches between x and y, all combinations of the matches are returned.

- `right_join(x, y, by = )`: return all rows from y, and all columns from x and y. Rows in y with no match in x will have NA values in the new columns. If there are multiple matches between x and y, all combinations of the matches are returned.

- `inner_join(x, y, by = )`: return all rows from x where there are matching values in y, and all columns from x and y. If there are multiple matches between x and y, all combination of the matches are returned.

- `full_join(x, y, by = )`:return all rows and all columns from both x and y. Where there are not matching values, returns NA for the one missing.

How are these different?

------------------------------------------------------------------------

And again with some simple data. *This data is from the join chapter on stat545.com, linked later in the notes.*

At each step, note what variables and rows appear.

*Note:* I don't specify which key/identifier variable to use, so R finds variables with the same names and matches using them. (It prints out what it uses so I can make sure it guessed right.)

```{r}
# make tibbles of data manually

superheroes <- tibble::tribble(
       ~name, ~alignment,  ~gender,          ~publisher,
   "Magneto",      "bad",   "male",            "Marvel",
     "Storm",     "good", "female",            "Marvel",
  "Mystique",      "bad", "female",            "Marvel",
    "Batman",     "good",   "male",                "DC",
     "Joker",      "bad",   "male",                "DC",
  "Catwoman",      "bad", "female",                "DC",
   "Hellboy",     "good",   "male", "Dark Horse Comics"
  )


publishers <- tibble::tribble(
  ~publisher, ~yr_founded,
        "DC",       1934L,
    "Marvel",       1939L,
     "Image",       1992L
  )

superheroes
publishers
```

------------------------------------------------------------------------

Inner join: Return all rows from x where there are matching values in y, and all columns from x and y.

What is missing? Why?

```{r, message = F}
inner_join(superheroes, publishers)
```

Think: how will `inner_join(publishers, superheroes)` be different?

. . .

```{r}
inner_join(publishers, superheroes)
```

------------------------------------------------------------------------

Left join: Return all rows from x, and all columns from x and y. If there are multiple matches between x and y, all combination of the matches are returned.

```{r}
left_join(superheroes, publishers)
```

------------------------------------------------------------------------

Right join: Return all rows from y, and all columns from x and y. If there are multiple matches between x and y, all combination of the matches are returned.

*Note:* Not often used. This is the same as `left_join(publishers, superheroes)`

```{r}
right_join(superheroes, publishers)
```

------------------------------------------------------------------------

Full join: Return all rows and all columns from both x and y. Where there are not matching values, returns NA for the one missing

```{r}
full_join(superheroes, publishers)
```

------------------------------------------------------------------------

### Filtering Joins

Filtering joins keep just the cases from the first data frame, but filters using the second.

- `semi_join(x, y, by = )`: return all rows from x where there are matching values in y, keeping just columns from x.

- `anti_join(x, y, by = )`: return all rows from x where there are not matching values in y, keeping just columns from x.

------------------------------------------------------------------------

Semi join: Return all rows from x where there are matching values in y, keeping just columns from x.

```{r}
semi_join(superheroes, publishers)
```

How is this *different* than the result from `inner_join(superheroes, publishers)`?

. . .

How will this compare to `semi_join(publishers, superheroes)`?

. . .

```{r}
semi_join(publishers, superheroes)
```

------------------------------------------------------------------------

Anti join: Return all rows from x where there are not matching values in y, keeping just columns from x

```{r}
anti_join(superheroes, publishers)
```

I also suggest you read through/reference:

- Chapter 11 of our textbook

- Chapter 19 of R for Data Science: <https://r4ds.hadley.nz/joins.html>

- this lecture from Jenny Bryan: <https://stat545.com/join-cheatsheet.html>

------------------------------------------------------------------------

## Examples

```{r}
#| include: false
donations <- tribble(
  ~Name,     ~Amount, ~Date,
  "Sam",     100,     "2/12/2019",
  "Mike",    50,      "2/12/2019",
  "Alice",   100,     "4/23/2019",
  "Mike",    50,      "5/17/2019",
  "Maria",   100,     "6/7/2019",
  "Susanne", 50,      "7/2/2019"
) |>
  dplyr::mutate(Date = mdy(Date))

donors <- tribble(
  ~Name,     ~Email,
  "Sam",     "sam@gmail.com",
  "Alice",   "aliceemail@gmail.com",
  "Susanne", "susanneroks@gmail.com",
  "Mike",    "michael@hotmail.com",
  "Mary",    "mary@gmail.com"
)
```

Consider our two tables from the introduction:

```{r}
donations
donors
```

1.  Let's write a join that would give us the information we need to send out an end of year receipt to our donors with their annual total.

```{r}
## full_join - this is giving EVERYTHING. Including, the email addresses for people that didn't donate. I don't need those. 
## left_join() - giving me every donation, along with the emails, when they exist. 
left_join(donations, donors, join_by(Name)) 

left_join(donations, donors, join_by(Name)) |> group_by(Name, Email) |> summarise(Total_donations = sum(Amount))
```

1.  Write a different join that emails the people in our database that haven't donated yet.

```{r}
anti_join(donors, donations)
```

```{r}
donations2 <- tribble(
  ~First_Name,     ~Amount, ~Date,
  "Sam",     100,     "2/12/2019",
  "Mike",    50,      "2/12/2019",
  "Alice",   100,     "4/23/2019",
  "Mike",    50,      "5/17/2019",
  "Maria",   100,     "6/7/2019",
  "Susanne", 50,      "7/2/2019"
) |>
  dplyr::mutate(Date = mdy(Date))

donors2 <- tribble(
  ~Name,     ~Email,
  "Sam",     "sam@gmail.com",
  "Alice",   "aliceemail@gmail.com",
  "Susanne", "susanneroks@gmail.com",
  "Mike",    "michael@hotmail.com",
  "Mary",    "mary@gmail.com"
)
```

```{r}
## the column names don't match anymore
donations2
donors2

left_join(donations2, donors2, join_by(First_Name == Name))
```

# Practice

Go to the in class lab for today.
